Dissecting the relationship between protein structure and sequence variation

### 2015-05-09 manuscript/cordiv/ej_manuscript_cordiv.tex
Creating a new version of the paper

\documentclass[11pt]{article}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{graphicx}
\usepackage{footnote}
\usepackage{rotating}
\usepackage{slashbox}
\usepackage{afterpage}
\usepackage{float}
\citestyle{apj}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%force setting figures at the top of the page:
    \makeatletter
    \setlength{\@fptop}{0pt}
    \makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  personal abbreviations and macros
%    the following package contains macros used in this document:
%\input econfmacros.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textwidth=6.9in  \textheight=9.6in
%%  Adjust these for your printer:
\leftmargin=-5in   \topmargin=-0.8in
\hoffset = -0.95in  \voffset = 0pt
\marginparwidth = 0pt
%% preprint number data:
%\newcommand\pubnumber{Article 14 in eConf C1304143}
\newcommand\pubnumber{}
\newcommand\pubdate{\today}

%%  address and funding acknowledgement data:
\def\affiliation{Department of Physics \\
The University of Texas at Austin, TX 78712, USA; amir@physics.utexas.edu}
%\def\support{\footnote{Work supported by the BEACON-NSF center for the study of evolution in action, under contract xxxxxx.}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   document style macros
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\gtrsim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$>$}}}}
\def\lessim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$<$}}}}
\newcommand{\ddg}{$\Delta\Delta G~$}
%\newcommand{\se}{\it seqent}
%\newcommand{\rs}{\it r4sJC}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\Title#1{\begin{center} {\Large #1 } \end{center}}
\def\Author#1{\begin{center}{ \sc #1} \end{center}}
\def\Address#1{\begin{center}{ \it #1} \end{center}}
\def\andauth{\begin{center}{and} \end{center}}
\def\submit#1{\begin{center}Submitted to {\sl #1} \end{center}}
\newcommand\pubblock{\rightline{\begin{tabular}{l} \pubnumber\\
         \pubdate  \end{tabular}}}
\newenvironment{Abstract}{\begin{quotation}  }{\end{quotation}}
%\newenvironment{Presented}{\begin{quotation} \begin{center}
%             PRESENTED AT\end{center}\bigskip
%      \begin{center}\begin{large}}{\end{large}\end{center} \end{quotation}}
\def\Acknowledgements{\bigskip  \bigskip \begin{center} \begin{large}
             \bf ACKNOWLEDGEMENTS \end{large}\end{center}}

@@

\begin{document}
\begin{titlepage}
\pubblock

\vfill
%\Title{Sequence Divergence as the main modulator of structure-sequence correlations}
\Title{Dissecting the relationship between protein structure and sequence variation}
\vfill
\Author{Amir Shahmoradi, Eleisha Jackson, Claus Wilke}%\support}
\Address{\affiliation}
\vfill
\begin{Abstract}
    Over the past decade several independent works have shown that some structural properties of proteins are capable of predicting protein evolution. The strength and significance of these structure-sequence relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. Here we present the results from a comprehensive search for the potential biophysical and structural determinants of protein evolution by studying more than $200$ structural and evolutionary properties in a dataset of $209$ monomeric enzymes. We discuss the main protein characteristics responsible for the general patterns of protein evolution, and identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in sequence-structure relations. In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations. In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest sequence-structure relations.

    %Over the past decade several independent works have shown that some site-specific structural properties of proteins are capable of predicting the site-specific evolutionary sequence variations. The strength and significance of these structure-evolution relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. In search for potential determinants of the observed variation in sequence-structure correlations, here we present the results of a comprehensive study of interrelations among more than $200$ structural and evolutionary characteristics in a dataset of 209 monomeric enzymes. We identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in correlation strengths. In addition to sequence divergence, we identify several protein structural properties that are weakly but significantly coupled with the strength of sequence-structure relations. In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to exhibit the strongest sequence-structure relations.
\end{Abstract}
\vfill
\vfill
\end{titlepage}
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}
%

@@

    Patterns of amino acid sequence variation are known to be influenced by the function of proteins (xx). The general consensus, based on the flurry of research done over the past several decades, is that the amino-acid sequence determines the 3D structure of proteins, known as the native conformation. This sequence-structure relation, however, does not necessitate a unique one-to-one mapping of sequence and the functionality of the protein. According to stability threshold model of proteins (xx), some substitutions at specific sites may be tolerated, if the new amino acid does not significantly change the energy landscape of the protein and therefore, its functional {\it native} conformation. Indeed, several independent work have identified site-specific structural properties that can explain the general patterns of sequence variability in proteins (xx). One of the earliest discovered examples of such relations, is the correlation of the site-specific Relative Solvent Accessibility (RSA) with sequence variation, that is, residues that are buried in the core of proteins tend to be more evolutionary conserved than exposed residues close to the surface of the protein.
    \\

    Other structural properties have also been recently identified and proposed to influence or explain the site-specific evolutionary variations of proteins. Among the simplest properties is the residue {\it contact number (CN)}, a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest (xx). Variants of this quantity that eliminate the free-parameter (i.e., the radius of the spherical neighborhood) in the definition of CN have also been proposed (xx) and have been shown to correlate better with sequence evolutionary rates (xx).   In a recent work Echave et al. xx presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg) to the rate at which mutations accumulate in the corresponding site over evolutionary time. They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
    \\

    Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different structures. By analyzing a data set of $216$ monomeric enzymes, Yeh et al. xx found a wide range of $\rho\sim0.1-0.8$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and the residue-specific solvent accessibility (i.e., RSA). Similarly, Echave et al. xx recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates. It appears that sequence-structure correlations tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}. This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
    \\

    The fact that all relevant structural properties seem to have more or less the same predictive power for sequence variability, implies the existence of one or more structural or evolutionary characteristics of proteins that modulate sequence-structure correlations in all proteins. Motivated by these observations, here we present the results of comprehensive effort in search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.  We show that among all properties considered, sequence divergence appears to be the primary determinant for the strength of sequence-structure relations. In addition, we show that proteins with more homogeneous Hydrogen bond (H-bond) energies, higher fraction of Helical structures and lower number of $\beta$-sheets generally tend to exhibit the strongest sequence-structure correlations. In the following sections, we present evidence in support of these findings and discuss their implications.
    \\

@@

    \subsection*{Sequence Data, Alignments and Evolutionary Rates}

    The results presented in this work are based on a dataset of $209$ monomeric enzymes (Echave papers, Wilke ddg paper xx) randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al. 2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes (Webb 1992) and domains of all main SCOP structural classes (Murzin et al. 1995). To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by (Yeh et all xx) for each protein from the {\it Clean Uniprot} database following the ConSurf protocol (Goldenberg et al. 2009; Ashkenazy et al. 2010). Sequence alignments were then constructed using amino-acid sequences with MAFFT (Katoh et al. 2002, 2005), specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data. The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset. To do so, we relied on two independent methods of measuring sequence variability measure. First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:

@@

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. Alternatively, we also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software rate4site (xx). To do so, first the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity (Stamatakis 2014). For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) (Mayrose et al. 2004).

@@

    The goal of the presented work is to identify the prominent structural or evolutionary properties of proteins that modulate sequence-structure correlations. These potential modulators represent a unique characteristics of the protein as a whole.  In general, the structural and evolutionary properties fall into two major categories.  1. {\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence. Prominent examples of the site-specific structural properties include RSA (Tien et al. 2012 xx), WCN (shih? xx).   2. {\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole. Examples include pdb Contact Order (CO) as defined by xx, protein size and compactness, sequence length, structural resolution of the protein in X-ray crystallography.  In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein. Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates. A comprehensive list of protein properties and their definitions are given in Table \ref{tab:definitions}.

@@

    In order to identify the potential determinants of sequence-structure correlations, we first ran a comprehensive search to identify site-specific structural properties that might correlate with measures of sequence variability (i.e., seqent \& r4s). There are however degeneracies in the definition of the some site-specific variables. For example, the quantity WCN is generally calculated from the coordinates of $\alpha$-carbon atoms in the 3-dimensional structure of proteins. There is however no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins. The same definition degeneracy also exists for the set of atomic Bfactors (xx) that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic Bfactor (e.g., Halle 2001 xx).
    \\

@@

    We use VORO++ software (xx) to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP (xx) for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al (20112 xx) to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins. In addition to ASA values, we also extract from DSSP output, information about the secondary structure of proteins such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns. To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structure and sequence properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations.
    \\

    All data including a list of 209 proteins and their properties together with Python, R and Fortran codes written for data reduction and analysis are publicly available to view and download at \url{https://github.com/shahmoradi/cordiv}.

@@

        To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also use multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the regression models, and the set of pdb-level properties as the explanatory variables.  Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression (reference to R package xx) on the entire dataset, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data. Depending on value of the free parameter $\alpha$, this generalized regression model is a compromise between {\it ridge regression} -- which attempts to shrink the coefficients of correlated predictors towards each other -- and {\it lasso regression} -- which tends to pick one of the correlated predictors and discard the rest. In addition to regularized regression, we have also employed Principal Component Regression (PCR) on the original dataset and its rank transformation. Both regression methods, PCR \& regularized, point to similar set of pdb-level properties as the strongest determinants of sequence-structure correlations.

\section{Discussion and Concluding Remarks}
\label{sec:dcr}

@@

\Acknowledgements

....

@@

%\section*{Figures}

    \begin{figure}[tbh]
            \begin{center}
            %\includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
            \includegraphics[width=6.9in]{cordiv_similarities_seqent.pdf}
            \end{center}
            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs. correlations of other structural properties with evolutionary rates. Detailed description of the structural properties is given in Section \label{sec:mam}. The red lines in each plot represent equality line. It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures. For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
            \label{fig:seqent_structure_cors}
    \end{figure}

%    \begin{figure}[tbh]
%        \begin{center}
%        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_two_in_one.pdf}
%        \end{center}
%        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset. In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other. The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN. The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
%        \label{fig:best_wcn}
%    \end{figure}

@@

    \begin{figure*}[tbh]
        \centering
        %\begin{tabular}{cc}
        \includegraphics[width=3.4in]{/validation_r_r4s_wcn_var_seqent.pdf}
        %\end{tabular}
        \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy. The black circles represent $209$ proteins used in this work. For comparison and validation, the red circles represent data for $9$ viral proteins taken from {shahmoradi2014xx}}.
        \label{fig:validation}
    \end{figure*}

%\newpage

%\section*{Supplementary Figures}

    %\centerline{\includegraphics[width=6.5in]{PC_screen_entropy_CSrmsf.pdf}}
    %\noindent \textbf{Fig. S1} Principal Component (PC) Regression of sequence entropy against the structural variables, including CS RMSF. {\bf (A)} Variance in entropy explained by each principal component. For most proteins, PC1 and either PC2 or PC3 show the strongest correlations with sequence entropy. Significant correlations ($P<0.05$) are shown as filled symbols, and insignificant correlations ($P\geq0.05$) are shown as open symbols. {\bf (B)} and {\bf (C)} Composition of the three leading components. Red arrows represent the loadings of each of the structural variables on the principal components; black dots represent the amino acid sites in the PC coordinate system.
    %\customlabel{fig:cor_entropy_PC_screen_CSrmsf}{S1}

%    \begin{figure}[tbh]
%        \begin{center}
%        \includegraphics[width=6.9in]{/best_wcn/selected_variables/boxplot_wcn_all_in_one.pdf}
%        \end{center}
%        \caption{A comparison of the predictive power of different measures of Weighted Contact Number (WCN) about different structure or sequence properties for 209 proteins in dataset. In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of WCN on the horizontal axis are compared against each other. The capital letters in the variable names on the horizontal axis denote the set of atomic coordinates used to calculate WCN. The variable $wcnSC$ denotes WCN measure based on the average coordinates over all heavy atoms in the side chain and the variable $wcnAA$ denotes WCN measure based on the average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
%        \label{fig:best_wcn}
%    \end{figure}
%
%    \begin{figure}[tbh]
%        \begin{center}
%        \includegraphics[width=6.9in]{/best_bf/selected_variables/boxplot_bf_all_in_one.pdf}
%        \end{center}
%        \caption{A comparison of the predictive power of different measures of Bfactor (BF) with different structure or sequence properties for 209 proteins in dataset. In each plot, the Spearman correlation strengths of a given structure or sequence property on the vertical axis with different measures of atomic Bfactors on the horizontal axis are compared against each other. The capital letters in the variable names on the horizontal axis denote the representative atomic Bfactors. The variable $bfSC$ denotes the average coordinate over all heavy atoms in the side chain for each site in protein and the variable $bfAA$ denotes average coordinate over all heavy atoms in the side chain and backbone of the amino acid.}
%        \label{fig:best_bf}
%    \end{figure}
%
%    \begin{figure}[tbh]
%        \begin{center}
%        \includegraphics[width=6.9in]{/best_varea/selected_variables/boxplot_varea_all_in_one.pdf}
%        \end{center}
%        \caption{Should be likely modified to volume-area-length-sphericity measures instead.}
%        \label{fig:best_voronoi}
%    \end{figure}

>>>>

\documentclass[12pt]{article}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{nicefrac}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{graphicx}
\usepackage{footnote}
\usepackage{rotating}
%\usepackage{slashbox}
\usepackage{afterpage}
\usepackage{float}
\usepackage{color}
%\citestyle{apj}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%force setting figures at the top of the page:
    \makeatletter
    \setlength{\@fptop}{0pt}
    \makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  personal abbreviations and macros
%    the following package contains macros used in this document:
%\input econfmacros.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textwidth=6.9in  \textheight=9.6in
%%  Adjust these for your printer:
\leftmargin=-5in   \topmargin=-0.8in
\hoffset = -0.95in  \voffset = 0pt
\marginparwidth = 0pt
%% preprint number data:
%\newcommand\pubnumber{Article 14 in eConf C1304143}
\newcommand\pubnumber{}
\newcommand\pubdate{\today}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   document style macros
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def\gtrsim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$>$}}}}
\def\lessim{\mathrel{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$<$}}}}
\newcommand{\ddg}{$\Delta\Delta G~$}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{Dissecting the relationship between protein structure and sequence variation}
\author{Amir Shahmoradi$^1$, Eleisha L. Jackson$^2$, Claus O. Wilke$^2$}%\support}
\begin{document}

\date{\today}
\maketitle

\noindent
$^1$ Department of Physics, The University of Texas at Austin, TX 78712, USA \\
$^2$ Institute of Cellular and Molecular Biology, Center for Computational Biology and Bioinformatics, and Department of Integrative Biology, The University of Texas at Austin, Austin, Texas, USA\\

\bigskip
\noindent
$^*$Corresponding author\\
$\phantom{^*}$Email: amir@physics.utexas.edu\\
$\phantom{^*}$Phone:{ \color{red} Need a phone} \\
%$\phantom{^*}$Phone: +1 512 232 2459\\

\bigskip
\noindent
Manuscript type: research article\\
\bigskip
\noindent  Keywords: protein evolution, relative solvent accessibility, site variability

\begin{abstract}
    Over the past decade several independent works have shown that some structural properties of proteins are capable of predicting protein evolution. The strength and significance of these structure-sequence relations, however, appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. Here we present the results from a comprehensive search for the potential biophysical and structural determinants of protein evolution by studying more than $200$ structural and evolutionary properties in a dataset of $209$ monomeric enzymes. We discuss the main protein characteristics responsible for the general patterns of protein evolution, and identify sequence divergence as the main determinant of the strengths of virtually all structure-evolution relationships, explaining $\sim10-30\%$ of observed variation in sequence-structure relations. In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations. In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest sequence-structure relations.

\end{abstract}
\vfill
\vfill
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}
%

@@

    Patterns of amino acid sequence variation are known to be influenced by the function of proteins (xx). The general consensus, based on the flurry of research done over the past several decades, is that the amino-acid sequence determines the 3D structure of proteins, known as the native conformation. This sequence-structure relation, however, does not necessitate a unique one-to-one mapping of sequence and the functionality of the protein. According to stability threshold model of proteins {\color{red}(xx)}, some substitutions at specific sites may be tolerated, if the new amino acid does not significantly change the energy landscape of the protein and therefore, its functional {\it native} conformation. Indeed, several independent work have identified site-specific structural properties that can explain the general patterns of sequence variability in proteins {\color{red}(xx)}. One of the earliest discovered examples of such relations, is the correlation of the site-specific Relative Solvent Accessibility (RSA) with sequence variation, that is, residues that are buried in the core of proteins tend to be more evolutionary conserved than exposed residues close to the surface of the protein.
    \\

    Other structural properties have also been recently identified and proposed to influence or explain the site-specific evolutionary variations of proteins. Among the simplest properties is the residue {\it contact number (CN)}, a measure of local density of the protein defined as the number of amino acids within a spherical neighborhood of a specific residue of interest {\color{red}(xx)}. Variants of this quantity that eliminate the free-parameter (i.e., the radius of the spherical neighborhood) in the definition of CN have also been proposed {\color{red}(xx)} and have been shown to correlate better with sequence evolutionary rates {\color{red}(xx)}.   In a recent work {\color{red}Echave et al. xx} presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg) to the rate at which mutations accumulate in the corresponding site over evolutionary time. They find that the variations in the free energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number.
    \\

    Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different structures. By analyzing a data set of $216$ monomeric enzymes, {\color{red}Yeh et al. xx} found a wide range of $\rho\sim0.1-0.8$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and the residue-specific solvent accessibility (i.e., RSA). Similarly, {\color{red}Echave et al. xx} recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates. It appears that sequence-structure correlations tend to correlate strongly with each other, as illustrated in Figure \ref{fig:ssscor}. This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates.
    \\

    The fact that all relevant structural properties seem to have more or less the same predictive power for sequence variability, implies the existence of one or more structural or evolutionary characteristics of proteins that modulate sequence-structure correlations in all proteins. Motivated by these observations, here we present the results of comprehensive effort in search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.  We show that among all properties considered, sequence divergence appears to be the primary determinant for the strength of sequence-structure relations. In addition, we show that proteins with more homogeneous Hydrogen bond (H-bond) energies, higher fraction of Helical structures and lower number of $\beta$-sheets generally tend to exhibit the strongest sequence-structure correlations. In the following sections, we present evidence in support of these findings and discuss their implications. %
    \\

@@

    \subsection*{Sequence Data, Alignments and Evolutionary Rates} {\color{red} Change this!!!!}

    The results presented in this work are based on a dataset of $209$ monomeric enzymes {\color{red}(Echave papers, Wilke ddg paper xx)} randomly picked from the Catalytic Site Atlas $2.2.11$ (Porter et al. 2004) with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes {\color{red}(Webb 1992)} and domains of all main SCOP structural classes (Murzin et al. 1995). To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by {\color{red}(Yeh et all xx)} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol {\color{red}(Goldenberg et al. 2009; Ashkenazy et al. 2010)}. Sequence alignments were then constructed using amino-acid sequences with MAFFT {\color{red}(Katoh et al. 2002, 2005)}, specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data. The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset. To do so, we relied on two independent methods of measuring sequence variability measure. First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:

@@

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. Alternatively, we also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software Rate4site {\color{red}(xx)}. To do so, first the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity {\color{red}(Stamatakis 2014)}. For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) {\color{red}(Mayrose et al. 2004)}.

@@

    The goal of the presented work is to identify the prominent structural or evolutionary properties of proteins that modulate sequence-structure correlations. These potential modulators represent a unique characteristics of the protein as a whole.  In general, the structural and evolutionary properties fall into two major categories.  1. {\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence. Prominent examples of the site-specific structural properties include RSA {\color{red}(Tien et al. 2012 xx),} WCN {\color{red}(shih? xx)}.   2. {\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole. Examples include pdb Contact Order (CO) as defined by {\color{red}xx}, protein size and compactness, sequence length, structural resolution of the protein in X-ray crystallography.  In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein. Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates. A comprehensive list of protein properties and their definitions are given in Table \ref{tab:definitions}.

@@

    In order to identify the potential determinants of sequence-structure correlations, we first ran a comprehensive search to identify site-specific structural properties that might correlate with measures of sequence variability (i.e., seqent \& r4s). There are however degeneracies in the definition of the some site-specific variables. For example, the quantity WCN is generally calculated from the coordinates of $\alpha$-carbon atoms in the 3-dimensional structure of proteins. There is however no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins. The same definition degeneracy also exists for the set of atomic Bfactors {\color{red}(xx)} that are used to represent site-specific flexibility, although the popular choice of residue flexibility is $\alpha$-carbon atomic Bfactor {\color{red}(e.g., Halle 2001 xx)}.
    \\

@@

    We use VORO++ software {\color{red}(xx)} to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP {\color{red}(xx)} for the calculation of Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of Tein et al {\color{red}(20112 xx)} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins. In addition to ASA values, we also extract from DSSP output, information about the secondary structure of proteins such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns. To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structure and sequence properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations.
    \\

All data and analysis scripts required to reproduce the work are publicly available are to view and download at \url{https://github.com/shahmoradi/cordiv}.

@@

        To ensure the accuracy of the results obtained from the Spearman correlation matrix of the pdb-level properties, we also use multivariate linear regression models, with individual sequence-structure correlations as the sole regressand of the regression models, and the set of pdb-level properties as the explanatory variables.  Since the number of explanatory variables is comparable to the number of observations (i.e., the number pdb structures in the dataset), we use regularized regression {\color{red}(reference to R package xx)} on the entire dataset, and also on the rank transformation of the dataset in order to minimize the effects of potential nonlinearities in data. Depending on value of the free parameter $\alpha$, this generalized regression model is a compromise between {\it ridge regression} -- which attempts to shrink the coefficients of correlated predictors towards each other -- and {\it lasso regression} -- which tends to pick one of the correlated predictors and discard the rest. In addition to regularized regression, we have also employed Principal Component Regression (PCR) on the original dataset and its rank transformation. Both regression methods, PCR \& regularized, point to similar set of pdb-level properties as the strongest determinants of sequence-structure correlations.

\section{Discussion}
\label{sec:dcr}

@@

\section{Acknowledgements}
The authors acknowledge the Texas Advanced Computing Center (TACC) at The University of Texas at Austin for providing High-performance computing resources. ELJ is is funded by a National Science Graduate Research Fellowship, grant number DGE-1110007. COW is funded by {\color{red} Which grants??}.  AS is funded by {\color{red} Which grants??}.
....

@@

%\cleardoublepage

\bibliographystyle{peerj} %"style
\bibliography{ej_manuscript_bib} %expected file "my refs.bib"

%\citep{Bloometal2006}
\section*{Figures}
    \begin{figure}[tbh]
            \begin{center}
            %\includegraphics[width=6.9in]{cordiv_similarities_r4sJC.pdf}
            \includegraphics[width=6.9in]{cordiv_similarities_seqent.pdf}
            \end{center}
            \caption{A comparison of the strength of Spearman correlation of sequence evolutionary rates (r4sJC) with {\it side chain} Weighted Contact Number vs. correlations of other structural properties with evolutionary rates. Detailed description of the structural properties is given in Section \label{sec:mam}. The red lines in each plot represent equality line. It is evident from all plots that for any given protein in dataset, the correlation strength of one structural property is a good proxy measure of the correlation strength of any other structural property with sequence variability measures. For brevity, correlations of structure-rate4site are not shown here but are available online, also in supplementary material.}
            \label{fig:seqent_structure_cors}
    \end{figure}

@@

    \begin{figure*}[tbh]
        \centering
        %\begin{tabular}{cc}
        \includegraphics[width=3.4in]{/validation_r_r4s_wcn_var_seqent.pdf}
        %\end{tabular}
        \caption{{\bf Sequence--structure correlation strength versus sequence divergence.} The plot illustrates the relationship between the strength of a representative sequence--structure correlation ({\it seqent--scnSC}) and the sequence divergence as measured by the variance of protein sequence entropy. The black circles represent $209$ proteins used in this work. For comparison and validation, the red circles represent data for $9$ viral proteins taken from {\color{red}shahmoradi2014xx}}.
        \label{fig:validation}
    \end{figure*}
